DisReserv (v1.0): Analytical displacement solution due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes

### 2020-09-18 manuscript/manuscript-reservoir.tex
theory and beginning of methodology

%%\twocolumn
\begin{abstract}
We have presented analytical solutions for the displacement due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes.
These solutions are based on the similarity between the gravitational potential yielded 
by a volume source under a density variation and the thermoelastic displacement 
potential yielded by a volume source in a half-space under a pressure variation.
This similarity enables the use of closed expressions of the gravitational potential 
and its derivatives for calculating the displacement and stress fields due to a volume 
source under a pressure variation.
We discretized the reservoir as a grid of 3D right rectangular prisms  juxtaposed in 
the horizontal and vertical directions.
Each prism has homogeneous pressure; however, pressure variations among different 
prisms are allowed. 
This parametrization of the reservoir yields a piecewise-constant distribution of 
pressure in the subsurface. 
The discrete reservoir modeling to calculate the displacement and stress fields due to 
this pressure variation follows the nucleus-of-strain concept, in which the center of 
each prism represents the coordinate of a nucleus of strain.
The nucleus of strain is considered an infinitesimal reservoir volume element.
The displacement and stress fields due each prism are calculated by integrating the 
nucleus of strain over the volume of the prism.
Finally, the displacement and stress due to the reservoir are computed by the sum of 
the contributions of each prism.
We provide a Python implementation (DisReserv) of our method to calculate the 
displacement and stress fields due to a reservoir with arbitrary shape and distribution 
of pressure changes. 
Our expressions are valid both inside and outside the reservoir. 
We calculate the displacement and stress fields in three scenarios: i) cylindrical 
reservoir under uniform depletion; ii) cylindrical reservoir under non uniform
depletion and iii) reservoir with arbitrary geometry and under arbitrary pressure 
distribution.
The results obtained for the three scenarios show that the Beltrami's and equilibrium 
equations are satisfied, which means that our method produces null stress field 
components at the free surface.
\end{abstract}

@@

The subsidence or displacement, stress and strain fields in the subsurface caused by reservoir compactation due to hydrocarbon production are grounded on the theory of thermoelasticity.

The Goodier’s thermoelastic displacement potential $\phi$ satisfies the Poisson's equation \citep{Goodier37}, i.e.:
\begin{equation}
\nabla^{2} \phi =  m \: T,
\label{eq:poisson}
\end{equation}
where $\nabla^{2}$ is the Laplacian operator, $T $ is the temperature difference and 
\begin{equation}
m =  \:  \alpha \: \frac{1 + \nu}{ 1 -\nu},
\label{eq:m}
\end{equation}
where $\alpha$  is the coefficient of linear thermal expansion and $\nu$ is the Poisson's ratio.

From the potential theory, a particular  solution of equation \ref{eq:poisson} is
\begin{equation}
\phi(x,y,z) = -  \frac{m}{4 \pi} \int\int\limits_{v}\int \frac{T(x^{\prime}, y^{\prime}, z^{\prime} )}{\sqrt{(x - x^{\prime})^{2} + (y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv,
\label{eq:phi}
\end{equation}
where $\phi(x,y,z)$ represents the Newtonian gravitational potential \citep{Kellogg29} at the coordinates $x, y$ and $z$ produced by a continuous distribution of mass defined by a set of very small masses $ -  \frac{m}{4 \pi}  \: T( x^{\prime}, y^{\prime}, z^{\prime} ) \: dv$, where $T(x^{\prime}, y^{\prime}, z^{\prime}) $ stands for the density distribution. 
The integral in equation \ref{eq:phi} is conducted over the coordinates $x^{\prime}$,$y^{\prime}$ and $z^{\prime}$, denoting, respectively, the $x-$, $y-$ and $-z$ coordinates of an arbitrary point belonging to the interior of the volume $v$ of the gravity source.

From equation \ref{eq:phi} and the potential-field theory, \citet{Goodier37} showed that if an element of volume $dv$ in the infinite solid is at a temperature $T(x^{\prime}, y^{\prime}, z^{\prime}) $, the remainder being at temperature zero, the displacement vector $\bf{u}$ caused by this temperature is the gradient of the Goodier’s thermoelastic displacement potential, i.e.,
\begin{equation}
{\bf{u}} = {\bf {\nabla}} \phi(x,y,z). 
\label{eq:displacement-general}
\end{equation}
where $\bf{\nabla}$ is the gradient operator.

To a homogenous half-space, \citet{Mindlin&Cheng50} showed that the method proposed by  \citet{Goodier37}  can be extended by the displacement solution given by:
\begin{equation}
\bf{u} = {\bf{\nabla}} \: \phi_{1} \: + \: {\bf{{\nabla}_{2}}} \phi_{2}, 
\label{eq:displacement-solution}
\end{equation}
where $\phi_{1} \equiv \phi_{1}(x,y,z)$ is the potential defined in equation \ref{eq:phi}, 
$\phi_{2} \equiv  \phi_{2}(x,y,z)$
is defined as "image potential" \citep{Segall92} due to a image point at the coordinates
$(x^{\prime}, y^{\prime}, -z^{\prime} )$ and the operator $\bf{{\nabla}_{2}}$ is expressed by
\begin{equation}
{\bf{{\nabla}_{2}}} = (3 - 4\nu){\bf{\nabla}} + 2 {\bf{\nabla}} z \frac{\partial }{\partial z}  - 4(1- \nu){\bf{\hat{z}}} {\bf{{\nabla}^{2}_{z}}},
\label{eq:nabla2}
\end{equation}
where ${\bf{\hat{z}}}$ is the unit vector in the $z-$direction and 
${\bf{{\nabla}^{2}_{z}}}$ is a scalar operator in which the operand is firtly multiplied by z and then operated upon by Laplacian operator ${\bf{\nabla}^{2}}$.

Equation \ref{eq:displacement-solution} is the displacement solution for the variation of temperature due to a single nucleus of strain buried at depth $z^{\prime}$ in a
semi-infinite homogeneous medium. 
In the right hand side of equation \ref{eq:displacement-solution}, the first term 
$ {\bf{\nabla}} \: \phi_{1} $ represents the displacement in an infinite medium, and the second term represents a correction of the displacement due a half-space, also known as image nucleus solution.

@@

Let's assume that a reservoir in the interior of the Earth is subject to a compactation due to hydrocarbon production. 
The compactation is caused by the pressure change within the reservoir, which in turn causes  a surface subsidence (or surface displacement). 
We discretized the reservoir into an $ m_{x} \times m_{y} \times m_{z} $ grid of 3D vertical juxtaposed prisms $(m_{x} \cdot m_{y} \cdot m_{z} = M)$ in which
the pressure within each prism is assumed to be constant and known. 
Each grid prism in the reservoir model may undergo a distinct pressure change.
Hence, the subsidence effect is the displacement field due to the pressure change throughout the reservoir and it can be calculated by the sum of the displacement produced by each prism.

The discrete forward modeling to calculate the displacement and stress fields due to a piecewise-constant distribution of the pressure contrast within a reservoir follows the nucleus of strain approach. 
By assuming that the center of each prism represents the coordinate of a nucleus of
strain, we can calculate the displacement field due to the pressure contrast of this prism by integrating over its volume.
The displacement solution for a single nucleus of strain in a homogeneous elastic semi-infinite medium (equation \ref{eq:displacement-solution}) will be used as an element of the displacement.

@@

Here, we use a Cartesian coordinate system with the $x-$axis pointing to north, the $y-$axis pointing to east and the $z-$axis pointing downward.
By considering the discrete form of equation \ref{eq:displacement-solution}, the displacement field  $ {\bf{u}}_{i} \equiv  {\bf{u}}(x_{i}, y_{i}, z_{i})$ at an arbitrary point $(x_{i}, y_{i}, z_{i})$ due to the $j$th nucleus of strain at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ will be calculated by
\begin{equation}
{\bf{u}}_{i}  = {\bf{\nabla}} \: 
{\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \: + \: {\bf{{\nabla}_{2}}} {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}, 
\label{eq:displacement_ui}
\end{equation}
In equation \ref{eq:displacement_ui}, the functions 
$\phi_{1} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}$ and  
$\phi_{2} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, -z^{\prime}_{j})}$
are, respectively, given by 
\begin{equation}
\phi_{1} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_1}_{ij} }
\label{eq:nucleus-phi1}
\end{equation}
and
\begin{equation}
\phi_{2} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_2}_{ij} }.
\label{eq:nucleus-phi2}
\end{equation}

>>>>

%%\twocolumn
\begin{abstract}
We have presented analytical solutions for the displacement due to reservoir compaction with arbitrary geometry and under arbitrary pressure changes.
These solutions are based on the similarity between the gravitational potential yielded 
by a volume source under a density variation and the thermoelastic displacement 
potential yielded by a volume source in a half-space under a pressure variation.
This similarity enables the use of closed expressions of the gravitational potential 
and its derivatives for calculating the displacement and stress fields due to a volume 
source under a pressure variation.
We discretized the reservoir as a grid of 3D right rectangular prisms  juxtaposed in 
the horizontal and vertical directions.
Each prism has homogeneous pressure; however, pressure variations among different 
prisms are allowed. 
This parametrization of the reservoir yields a piecewise-constant distribution of 
pressure in the subsurface. 
The discrete reservoir modeling to calculate the displacement and stress fields due to 
this pressure variation follows the nucleus-of-strain concept.
The nucleus of strain is considered an infinitesimal reservoir volume element.
The displacement and stress fields due each prism are calculated by integrating the 
nucleus of strain over the volume of the prism.
Finally, the displacement and stress due to the reservoir are computed by the sum of 
the contributions of each prism.
We provide a Python implementation (DisReserv) of our method to calculate the 
displacement and stress fields due to a reservoir with arbitrary shape and distribution 
of pressure changes. 
Our expressions are valid both inside and outside the reservoir. 
We calculate the displacement and stress fields in three scenarios: i) cylindrical 
reservoir under uniform depletion; ii) cylindrical reservoir under non uniform
depletion and iii) reservoir with arbitrary geometry and under arbitrary pressure 
distribution.
The results obtained for the three scenarios show that the Beltrami's and equilibrium 
equations are satisfied, which means that our method produces null stress field 
components at the free surface.
\end{abstract}

@@

The displacement, stress and strain fields in the subsurface caused by reservoir compactation due to hydrocarbon production are grounded on the theory of thermoelasticity.

The Goodier’s thermoelastic displacement potential $\phi$ satisfies the Poisson's equation \citep{Goodier37}, i.e.:
\begin{equation}
\nabla^{2} \phi =  m \: T,
\label{eq:poisson}
\end{equation}
where $\nabla^{2}$ is the Laplacian operator, $T $ is the temperature variation and 
\begin{equation}
m =  \:  \alpha \: \frac{1 + \nu}{ 1 -\nu},
\label{eq:m}
\end{equation}
where $\alpha$ is the coefficient of linear thermal expansion and $\nu$ is the Poisson's ratio.

From the potential theory, a particular solution of equation \ref{eq:poisson} is
\begin{equation}
\phi(x,y,z) = -  \frac{m}{4 \pi} \int\int\limits_{v}\int \frac{T(x^{\prime}, y^{\prime}, z^{\prime} )}{\sqrt{(x - x^{\prime})^{2} + (y - y^{\prime})^{2} + (z - z^{\prime})^{2}}} \: dv,
\label{eq:phi}
\end{equation}
where $\phi(x,y,z)$ represents the Newtonian gravitational potential \citep{Kellogg29} that would be produced at the coordinates $x, y$ and $z$ by a continuous density distribution $- \frac{m}{4 \pi}  \: T( x^{\prime}, y^{\prime}, z^{\prime} )$. 
The integral in equation \ref{eq:phi} is conducted over the coordinates $x^{\prime}$,$y^{\prime}$ and $z^{\prime}$, denoting, respectively, the $x-$, $y-$ and $-z$ coordinates of an arbitrary point belonging to the interior of the volume $v$ of the solid.

From equation \ref{eq:phi} and the potential theory, \citet{Goodier37} showed that, if an element of volume $dv$ in the infinite solid is at a temperature $T(x^{\prime}, y^{\prime}, z^{\prime}) $, the remainder being at temperature zero, the displacement vector $\bf{u}$ caused by this temperature is the gradient of the Goodier’s thermoelastic displacement potential, i.e.,
\begin{equation}
{\bf{u}} = {\bf {\nabla}} \phi(x,y,z) \: ,
\label{eq:displacement-general}
\end{equation}
where $\bf{\nabla}$ is the gradient operator.

To a homogenous half-space, \citet{Mindlin&Cheng50} showed that the method proposed by  \citet{Goodier37} can be extended by the displacement solution given by:
\begin{equation}
\bf{u} = {\bf{\nabla}} \: \phi_{1} \: + \: {\bf{{\nabla}_{2}}} \phi_{2}, 
\label{eq:displacement-solution}
\end{equation}
where $\phi_{1} \equiv \phi_{1}(x,y,z)$ is the potential defined in equation \ref{eq:phi}, 
$\phi_{2} \equiv  \phi_{2}(x,y,z)$
is defined as "image potential" \citep{Segall92} due to a image point at the coordinates
$(x^{\prime}, y^{\prime}, -z^{\prime} )$ and the operator $\bf{{\nabla}_{2}}$ is expressed by
\begin{equation}
{\bf{{\nabla}_{2}}} = (3 - 4\nu){\bf{\nabla}} + 2 {\bf{\nabla}} z \frac{\partial }{\partial z}  - 4(1- \nu){\bf{\hat{z}}} {\bf{{\nabla}^{2}_{z}}},
\label{eq:nabla2}
\end{equation}
where ${\bf{\hat{z}}}$ is the unit vector in the $z-$direction and 
${\bf{{\nabla}^{2}_{z}}}$ is a scalar operator in which the operand is first to be  multiplied by $z$ and then operated upon by the Laplacian ${\bf{\nabla}^{2}}$.

Equation \ref{eq:displacement-solution} is the displacement solution for the variation of temperature due to a single nucleus of strain buried at depth $z^{\prime}$ in a
semi-infinite homogeneous medium. 
In the right hand side of equation \ref{eq:displacement-solution}, the first term 
$ {\bf{\nabla}} \: \phi_{1} $ represents the displacement in an infinite medium, and the second term represents a correction of the displacement due a half-space, also known as "image nucleus solution".

@@

Let's assume that a reservoir in the interior of the Earth is subject to a compactation due to hydrocarbon production. 
The compactation is caused by the pressure change within the reservoir, which in turn causes  a surface subsidence (or surface displacement). 
Here, we use a Cartesian coordinate system with the $x-$axis pointing to north, the $y-$axis pointing to east and the $z-$axis pointing downward.
We discretize the reservoir into an $ m_{x} \times m_{y} \times m_{z} $ grid of 3D vertical juxtaposed prisms $(m_{x} \cdot m_{y} \cdot m_{z} = M)$ along the 
$x$, $y$ and $z$ axes, respectively, in which the pressure within each prism is 
assumed to be constant and known. 
Each prism in the reservoir model may undergo a distinct pressure change.
The subsidence effect is the displacement field due to the pressure change throughout the reservoir and is calculated by the sum of the displacement produced by each prism.

The discrete forward modeling to calculate the displacement and stress fields due to a piecewise-constant distribution of the pressure contrast within a reservoir follows the nucleus-of-strain approach. 
We assume that a nucleus of strain represents an infinitesimal reservoir volume 
element. The displacement solution for a single nucleus of strain in a homogeneous elastic semi-infinite medium (equation \ref{eq:displacement-solution}) will be used as an element of the displacement.
Then we calculate the displacement field due to the pressure contrast of a prism by integrating the nucleus of strain over its volume.

@@

By considering the discrete form of equation \ref{eq:displacement-solution}, the displacement field  $ {\bf{u}}_{i} \equiv  {\bf{u}}(x_{i}, y_{i}, z_{i})$ at an arbitrary point $(x_{i}, y_{i}, z_{i})$ due to the $j$th nucleus of strain at the coordinates 
$(x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})$ will be calculated by
\begin{equation}
{\bf{u}}_{i}  = {\bf{\nabla}} \: 
{\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})} \: + \: {\bf{{\nabla}_{2}}} {\phi_{2}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}, 
\label{eq:displacement_ui}
\end{equation}
In equation \ref{eq:displacement_ui}, the functions 
$\phi_{1} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, z^{\prime}_{j})}$ and  
$\phi_{2} \equiv {\phi_{1}(x_{i}, y_{i}, z_{i}, x^{\prime}_{j}, y^{\prime}_{j}, -z^{\prime}_{j})}$
are, respectively, given by 
\begin{equation}
\phi_{1} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_1}_{ij} }
\label{eq:nucleus-phi1}
\end{equation}
and
\begin{equation}
\phi_{2} = - \frac{C_m}{4 \pi}  \frac{\Delta p_{j} \: \:dv_j}{ {R_2}_{ij} }.
\label{eq:nucleus-phi2}
\end{equation}
